---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

Alex Kodess
amkodess@gmail.com

http://www.cdc.gov/brfss/

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(tidyverse)
library(scales)
library(knitr)
```

### Load data
```{r load-data}
load("brfss2013.RData")
```



* * *

## Part 1: Data

**Generalizability**: 
The data appear to be collected from a fairly large group of people, in different locations 
around the United States, via a variety of methods. Hence conclusions drawn from the data 
are generalizable. 

**Causality**: 
However since the data are purely obsevational, all the conclusions are not causal. 
Only correlational relationships could be investigated.


* * *

## Part 2: Research questions

**Research quesion 1: **
Investigate the classical question about whether 
there exists a relationship between one's sleep times and 
one's health status. To avoid a possible occurence of 
Simpson's paradox, we divide observations by their gender.


**Research quesion 2:**
Investigate a possible relationship between one's 
self-reported health status and one's education level. Split 
the exploratory analysis by gender.

**Research quesion 3:**
How many people in New England cannot afford to see a doctor? 
We again split our analysis by gender.


* * *

## Part 3: Exploratory data analysis

NOTE: Insert code chunks as needed by clicking on the "Insert a new code chunk" 
button (green button with orange arrow) above. Make sure that your code is visible
in the project you submit. Delete this note when before you submit your work.

**Research quesion 1. Sleep and health: variables `genhtlh`, `sleptim1`, `sex`**

The variables relevant to this question are 
`genhlth`,
```{r}
str(brfss2013$genhlth); summary(brfss2013$genhlth)
```
`sleptim1`,
```{r}
str(brfss2013$sleptim1); summary(brfss2013$sleptim1)
```
and `sex`
```{r}
str(brfss2013$sex); summary(brfss2013$sex)
```


First we realize that there are missing data as well as data 
that are obviously incorrect such as sleeping times exceeding 
24 hours. We filter these out first.

```{r}
tb_1 <- brfss2013 %>% 
        select(genhlth, sleptim1, sex) %>% 
        filter(
                !is.na(genhlth), 
                !is.na(sleptim1), 
                !is.na(sex),
                sleptim1 <= 24
                )
```

We now find a summary of sleeping patterns among different groups 
of health status and different genders. The mean time is shown in minutes. 

```{r}
tb_1 %>% group_by(genhlth, sex) %>%
        summarise(med_sleep = median(sleptim1),
                  mean_sleep = round(60*mean(sleptim1))) %>%
        ungroup() %>%
        arrange(desc(mean_sleep)) 

```

These data are now visualized using side-by-side boxplots. The 
data are faceted by gender.

```{r}
ggplot(
        data = tb_1, 
        mapping = aes(x = genhlth, y = sleptim1)
        ) + 
        geom_boxplot() +
        scale_y_continuous(breaks = pretty_breaks(n = 24)) + 
        labs(x = "Health Status", y = "Sleep Time, hr") + 
        facet_wrap(~ sex)


```


**Research quesion 2. Education and health: variables  `genhlth`, `sex`, `educa`**

The variables relevant to this question are 
`genhlth`, `sex`, and `educa`
```{r}
str(brfss2013$educa); summary(brfss2013$educa)
```

We first obtan all the observations were all three variables 
are available.


```{r}
tb_2 <- brfss2013 %>% 
        select(genhlth, educa, sex) %>% 
        filter(
                !is.na(genhlth), 
                !is.na(educa), 
                !is.na(sex)
                )
```

We shorten the names of the factor levels of `educa` for 
aesthetic reasons.

```{r}
levels(tb_2$educa) <- c(
                                "Only kindergarten",
                                "Elementary",
                                "Some High School",
                                "High School graduate",
                                "College dropped",
                                "College graduate"
)
```
This gives us a contingency table for `educa` and `genhlth` 
(regardless of `sex`)
```{r}
cont_tab <- table(tb_2$educa, tb_2$genhlth)
round(100*prop.table(cont_tab), 4)
```

These data are now visualized using bar plots. The 
data are faceted by gender.

```{r}
ggplot(
        data = tb_2, 
        aes(x = educa, fill = genhlth)) + 
        geom_bar(position = "fill") + 
        labs(x = "Education Level", 
             y = "Proportion by Health Status"
        )+
        scale_fill_discrete('Health Status') + 
        coord_flip() + 
        facet_wrap(~ sex)
```



**Research quesion 3: How many people cannot afford a doctor in New England? variables `X_state`, `sex`, `medcost`**

We first of all realize that the variable `_state` is not present in the dataset, and 
we should use `X_state`. 
```{r}
str(brfss2013$X_state); summary(brfss2013$X_state)
```

The other two variables involved in this question are `sex` 
and `medcost`:
```{r}
str(brfss2013$medcost); summary(brfss2013$medcost)
```

We define `New_England` as the following six states in the Northest US.

```{r}
New_England <- c(
                        "Connecticut",
                        "Maine",
                        "Massachusetts",
                        "New Hampshire",
                        "Rhode Island",
                        "Vermont"
)
```
We select the needed variables, group them by gender and state, and 
compute in each group the proportion of people who answered positively 
to the question *Was there a time in the past 12 months when 
you needed to see a doctor but could not because of cost?*

```{r}
tb_3 <- brfss2013 %>% 
        select(X_state, medcost, sex) %>% 
        filter(
                X_state %in% New_England,
                !is.na(medcost),
                !is.na(sex)
        ) %>% 
        group_by(sex, X_state) %>%
        summarise(
                perc = length(medcost[medcost == "Yes"]) / n()
        )

print(tb_3)
```
These data are now visualized using bar plots. The data are faceted by gender.
```{r}
tb_3$X_state = factor(tb_3$X_state)
ggplot(
        data = tb_3, 
        aes(x = X_state, y = perc)) + 
        geom_col() + 
        labs(x = "State", y = "Percentage") + 
        coord_flip() + 
        facet_wrap(~ sex)
```